The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-11-21 paper.qmd
updated paper with correct spelling and sent to arxiv

---
title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
author: 
  - name: Harriet Mason
    url: https://harrietmason.netlify.app/
    orcid: 0009-0007-4568-8215
    email:  harriet.m.mason@gmail.com
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Dianne Cook
    url: https://dicook.org
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Sarah Goodwin
    url: https://www.linkedin.com/in/smgoodwin/
    orcid: 0000-0001-8894-8282
    email: Sarah.Goodwin@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Human Centred Computing
      city: Melbourne
      country: Australia
  - name: Emi Tanaka
    url: https://emitanaka.org/
    orcid: 0000-0002-1455-259X
    email: Emi.Tanaka@anu.edu.au
    affiliation: 
    - name: The Australian National University
      department: Biological Data Science Institute
      city: Canberra
      country: Australia
  - name: Susan VanderPlas
    url: https://srvanderplas.github.io
    orcid: 0000-0002-3803-0972
    email: susan.vanderplas@unl.edu
    affiliation:
    - name: University of Nebraska–Lincoln
      department: Statistics Department
      city:  Lincoln
      country: United States
bibliography: paper.bib
abstract: Uncertainty visualisation is quickly becomming a hot topic in information visualisation. Exisiting reviews in the field take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information. This conflict largely stems from a conflation between uncertainty visualisations designed for decision making and those designed to prevent false conclusions. We coin the term "signal suppression" to describe a visualisation that is designed for preventing false conclusions, as the approach demands that the signal (i.e. the collective take away of the estimates) is suppressed by the noise (i.e. the variance on those estimates). We argue that the current standards in visualisation suggest that uncertainty visualisations designed for decision making should not be considered uncertainty visualisations at all. Therefore, future work should focus on signal suppression. Effective signal suppression requires us to communicate the signal and the noise as a single "validity of signal" variable, and doing so proves to be difficult with current methods. We illustrate current approaches to uncertainty visualisation by showing how they would change the visual apprearance of a choropleth map. These maps allow us to see why some methods succeed at signal suppression, while others fall short. Evaluating visualisations on how well they perform signal suppression also proves to be difficult, as it involves measuring the effect of noise, a variable we typically try to ignore. We suggest authors use qualitative studies or compare uncertainty visualisations to the relevant hypothesis tests. 
date: last-modified
toc: false
number-sections: true
format:
  jasa-pdf:
    keep-tex: true  
    journal:
      blinded: false
  jasa-html: default
fig-valign: bottom
cap-location: bottom
editor_options: 
  chunk_output_type: console
---

>>>>

---
title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
author: 
  - name: Harriet Mason
    url: https://harrietmason.netlify.app/
    orcid: 0009-0007-4568-8215
    email:  harriet.m.mason@gmail.com
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Dianne Cook
    url: https://dicook.org
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Sarah Goodwin
    url: https://www.linkedin.com/in/smgoodwin/
    orcid: 0000-0001-8894-8282
    email: Sarah.Goodwin@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Human Centred Computing
      city: Melbourne
      country: Australia
  - name: Emi Tanaka
    url: https://emitanaka.org/
    orcid: 0000-0002-1455-259X
    email: Emi.Tanaka@anu.edu.au
    affiliation: 
    - name: The Australian National University
      department: Biological Data Science Institute
      city: Canberra
      country: Australia
  - name: Susan VanderPlas
    url: https://srvanderplas.github.io
    orcid: 0000-0002-3803-0972
    email: susan.vanderplas@unl.edu
    affiliation:
    - name: University of Nebraska–Lincoln
      department: Statistics Department
      city:  Lincoln
      country: United States
bibliography: paper.bib
abstract: Uncertainty visualisation is quickly becomming a hot topic in information visualisation. Exisiting reviews in the field take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information. This conflict largely stems from a conflation between uncertainty visualisations designed for decision making and those designed to prevent false conclusions. We coin the term "signal suppression" to describe a visualisation that is designed for preventing false conclusions, as the approach demands that the signal (i.e. the collective take away of the estimates) is suppressed by the noise (i.e. the variance on those estimates). We argue that the current standards in visualisation suggest that uncertainty visualisations designed for decision making should not be considered uncertainty visualisations at all. Therefore, future work should focus on signal suppression. Effective signal suppression requires us to communicate the signal and the noise as a single "validity of signal" variable, and doing so proves to be difficult with current methods. We illustrate current approaches to uncertainty visualisation by showing how they would change the visual apprearance of a choropleth map. These maps allow us to see why some methods succeed at signal suppression, while others fall short. Evaluating visualisations on how well they perform signal suppression also proves to be difficult, as it involves measuring the effect of noise, a variable we typically try to ignore. We suggest authors use qualitative studies or compare uncertainty visualisations to the relevant hypothesis tests. 
date: last-modified
toc: false
number-sections: true
latex-clean: true
format:
  jasa-pdf:
    keep-tex: true  
    journal:
      blinded: false
  jasa-html: default
fig-valign: bottom
cap-location: bottom
editor_options: 
  chunk_output_type: console
---
